---
layout: post
category: computation
tags:
- rfishbase
- ropensci
- R
---

```{r include=FALSE}
library(knitr)
opts_chunk$set(results='hide')
```


Using `dplyr` calls on the back-end of the `rfishbase` re-write means working around the non-standard evaluation (NSE), as described in the [`dplyr` vignette]().

Grab the data I was using for this:

```{r}
library("dplyr")
downloader::download("https://github.com/cboettig/2015/raw/fc0d9185659e7976927d0ec91981912537ac6018/assets/data/2015-02-06-taxa.csv", "taxa.csv")
all_taxa <- read.csv("taxa.csv")
```

Consider a simple NSE `dplyr` call:

```{r}
x <- filter(all_taxa, Family == 'Scaridae')
```

The best SE version of this just needs to use the formula expression, `~`, the `_` SE version of the function and it's `.dots` argument:

```{r}
.dots <- list(~Family == 'Scaridae')
x1 <- filter_(all_taxa, .dots=.dots)

identical(x, x1)
```

This lets us treat the arguments (e.g. values of the factor on which we filter) as variables:


```{r}
family <- 'Scaridae'
.dots <- list(~Family == family)
x2 <- filter_(all_taxa, .dots=.dots)
identical(x, x2)
```  

If we want both the key and value to vary, we need to get pretty fancy to subvert the non-standard evaluation:

```{r}
library(lazyeval)
family <- 'Scaridae'
field <- 'Family'
.dots <- list(interp(~y == x, 
                     .values = list(y = as.name(field), x = family)))
x3 <- filter_(all_taxa, .dots=.dots)
identical(x, x3)
```

At bit more fun to wrap this into a function where we take arbitrary number of arguments as name-value pairs:

```{r}
query <- list(Family = 'Scaridae', SpecCode = 5537)
dots <- lapply(names(query), function(level){
    value <- query[[level]]
    interp(~y == x, 
                .values = list(y = as.name(level), x = value))
  })
  
x3 <-  filter_(all_taxa, .dots = dots) 
```

```{r include=FALSE}
unlink("taxa.csv")
```


----

## More fun standardizing NSE

The previous examples show only applications to `filter_()`.  While the general idea is the same, this pattern doesn't translate directly for other functions, such as `mutate_`.  Here's some common patterns I've adopted when using `mutate_()`.  First consider the familiar NSE useage:

```{r}
df <- mutate(mtcars, displ_l = disp / 61.0237)
head(df)
```

Again we use `list(interp(` pattern, but note that we specify the name for our new column using `setNames` (naming the elements of the list).

```{r}
dots <- setNames(list(lazyeval::interp(~x / y, x = quote(disp), y=61.0237)), "displ_l")
df2 <- mutate_(mtcars, .dots = dots)
identical(df, df2)
```

Of course the use `y` could be skipped for a more direct value if that was not a variable.

## More `dplyr` patterns

Also thought I would scribble down some other common `dplyr` patterns I find myself re-using.

- applying a function that returns a `data.frame` to each element of a `list` and coercing the combined output to a `data.frame`:

```r
mylist %>% lapply(myfun) %>% dplyr::bind_rows() 
```

To place this deeper in the `hadleyverse`, `purrr::map` could be dropped in for `lapply` in the above example.  

- Another common pattern for me is `expand.grid() %>% group_by() %>% do()`, Here's a [recent example of mine](https://github.com/cboettig/multiple_uncertainty/blob/3f42aa4623f40cee70d5a9d52a8cce73617248a5/inst/scripts/table.Rmd)

Also includes an example of how to define `group_by_all()` since that is usually the grouping I need from an `expand.grid()` call (that is, I want to apply over all combinations of some parameter settings, etc)



Something I hope is not a common pattern but one I struggled with for a bit: making recursive calls of the above pattern for nested lists.  This code in RNeXML illustrates my solution, which required both [function recursion and function closure](https://github.com/ropensci/RNeXML/blob/b596839e3a1fc53e74b46fd935d91dca59a9da98/R/get_metadata.R#L50-L59).  

