---
layout: post
category: computing
tags: 
  - nimble
---




```{r}
library("nimble")
library("lazyeval")
```

Can we declare the `nimbleCode` parts in sections? Such as defining the model and priors in separate, reusable code blocks? Here's a "model" block:

```{r}
model <- nimbleCode({
  for (i in 1:N){
    theta[i] ~ dgamma(alpha,beta)
    lambda[i] <- theta[i]*t[i]
    x[i] ~ dpois(lambda[i])
  }
})
```


We may also wish to define the priors separately, and with specified list of hyperparameters for the priors.  Perhaps the need to use `quote` and to specify each line as a list item, instead of as a block defined with `{`, is not ideal, but non-standard evaluation is a bit of a wild west still.

```{r}
hyperparameters <- list(lambda = 1.0, a = 0.1, b = 1.0)

priors <- nimbleCode({
  alpha ~ dexp(lambda)
  beta ~ dgamma(a,b)
})

```

Having done so, we can construct a `nimbleCode` block by passing our chosen hyperparameters to the priors and concatenating the model and priors.  

```{r}
P <- as.expression(sapply(priors, lazyeval::interp, .values = hyperparameters)[-1])
pumpCode <- c(model, P)
```

The result appears to work as a functional `nimbleCode` block; though note the resulting expression is not quite identical to the one we would typically write (e.g. without commas):

```{r}
pumpCode
```

Here we just specify the rest of the model:

```{r}
pumpConsts <- list(N = 10, 
                   t = c(94.3, 15.7, 62.9, 126, 5.24,
                         31.4, 1.05, 1.05, 2.1, 10.5))
pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))
pumpInits <- list(alpha = 1, 
                  beta = 1,
                  theta = rep(0.1, pumpConsts$N))

pump <- nimbleModel(code = pumpCode, 
                    name = 'pump', 
                    constants = pumpConsts,
                    data = pumpData, 
                    inits = pumpInits)
```

and then verify that it runs.

```{r}
Cmodel <- compileNimble(pump)
mcmcspec <- configureMCMC(pump, print=FALSE)
mcmc <- buildMCMC(mcmcspec)
Cmcmc <- compileNimble(mcmc, project = Cmodel)
Cmcmc$run(1000)
samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
plot(samples[ , 'alpha'], samples[ , 'beta'], xlab = expression(alpha), ylab = expression(beta))
```

(The plot shows the same correlation issue that arises without using a block sampler that is illustrated in the manual.)
