---
layout: post

---

```{r message=FALSE, results='hide'}
library(FastGP)
library(mvtnorm)
library(MASS)
library(rbenchmark)
```


## Demo of elliptical slice sampling

Relevant Parameters:
```{r}
A <- 1 #amplitude of the sin function used for our signal
T <- 5 #period of the sin function used as our signal
sigma <- 10 #variance parameter in the covariance function
phi <- 100 #decay parameter for the exponential kernel
N <- 100 #number of time points 
```

A function that creates a signal, which can be toggled for multiple shapes. In the present iteration we use a sin function with variable period.

```{r}
signal <- function(t) {
  return(A*sin(t/T))
}
```


Define our covariance matrix using the exponential kernel

```{r}
S <- as.matrix(sigma*exp(-as.matrix(dist(seq(1,N)))^2/phi))
```

Generate a copy of the signal on the time scale of 1...N

```{r}
t <- mvrnorm(n = 1, mu = rep(0,100), Sigma = S , tol = 1e-6)
s <- signal(seq(1,N)+t)+rnorm(N,sd=sqrt(.001))
```

log-lik functions

````{r}
log_lik <- function(w,s){
  return (dmvnorm(s, mean = signal(seq(1:N)+w),sigma=.001*diag(1,N),log=T))
}

#rcpp based log-lik function
log_lik_rcpp <- function(w,s){
  return (rcpp_log_dmvnorm(S = .001*diag(1,N),mu=signal(seq(1:N)+w),s,istoep=TRUE))
}
```


```{r}
X <- matrix(seq(1,N),nrow=N)
Sig <- sigma*exp(-rcpp_distance(X,dim(X)[1],dim(X)[2])^2/phi)
```


and here we go:

```{r}
mcmc_samples <- ess(log_lik_rcpp,s,Sig,1000,250,100,TRUE)
```


```{r}
par(mfrow=c(1,1))
plot(colMeans(mcmc_samples), type="l")
points(t, type="l", lwd=3, col="red")
points(colMeans(mcmc_samples) + 2* apply(mcmc_samples, 2, sd), type="l", lty="dashed")
points(colMeans(mcmc_samples) - 2* apply(mcmc_samples, 2, sd), type="l", lty="dashed")
```


test `ess` with rcpp likelihood versus non rcpp likelihood

```{r}
benchmark(ess(log_lik_rcpp,s,Sig,1000,250,100,FALSE), ess(log_lik,s,Sig,1000,250,100,TRUE),replications=2)
```
