---
layout: post
category: ecology
tags: 
- gp
- nimble


---


### Setup

```{r message=FALSE}
devtools::install_github('nimble-dev/nimble/packages/nimble@devel')
library(nimble)
set.seed(0)
```

Randomly generate some sinusoidal data

```{r}
x <- 1:100
y <- sin(x/5) + rnorm(100, 0.1)
ind <- sort(sample(1:100, 40))
```

The next three lines are the only inputs required.

```{r}
xObs <- x[ind]                ## input
yObs <- y[ind]                ## input
xPred <- c(x, 101:120)        ## input
```

Some initial processing
```{r}
nObs <- length(xObs)
nPred <- length(xPred)
f <- function(xi, xj) (xi-xj)^2
diffOO <- outer(xObs,  xObs,  f)
diffPP <- outer(xPred, xPred, f)
diffPO <- outer(xPred, xObs,  f)
```



### NIMBLE function for GP prediction

Here's the main function for doing the prediction from the GP model.

It takes the MCMC samples as a *runtime* argument, so this can be iterated with running the MCMC for different numbers of iterations, initial values, or even different datasets!

```{r, warning=FALSE}
gpPred <- nimbleFunction(
    setup = function(model, params) {
        calcNodes <- model$getDependencies(params, determOnly = TRUE)
        nPred <- dim(model$SigPP)[1]
        E <- array(0, c(nPred, 1))
        C <- array(0, c(nPred, nPred))
    },
    run = function(samples = double(2)) {
        E <<- E * 0
        C <<- C * 0
        nSamples <- dim(samples)[1]
        for(i in 1:nSamples) {
            values(model, params) <<- samples[i,]
            calculate(model, calcNodes)
            intermediate <- model$SigPO %*% inverse(model$SigOO)
            Etemp <- intermediate %*% asCol(model$yObs)
            Ctemp <- model$SigPP - intermediate %*% t(model$SigPO)
            E <<- E + Etemp
            C <<- C + Ctemp
        }
        E <<- E / nSamples
        C <<- C / nSamples
    },
    methods = list(
        getE = function() { returnType(double(1)); return(E[,1]) },
        getC = function() { returnType(double(2)); return(C[, ]) }
    )
)
```


### GP model

GP model defined here.  Basically the same as your original, but I renamed some paramters more to my liking =)

This is *not* as elegant as I had hoped.  It still requires repeating (essentially) the same code three times.  I discovered some limitations of NIMBLE while trying other approaches.

Bottom line:  
- This achieves relatively nice simplicity.  
- There's an efficiency hit to the MCMC sampling, but it doesn't seem too bad.  
- This works.  

```{r, message=FALSE}
code <- nimbleCode({
   rho ~ dgamma(10, 1)
   sigGP ~ dunif(0, 1e5)
   sigOE ~ dunif(0, 1e5)
   SigOO[1:nObs, 1:nObs ] <- sigGP^2*exp(-1/2*diffOO[1:nObs, 1:nObs ]/rho^2) + sigOE^2*IOO[1:nObs, 1:nObs ]
   SigPP[1:nPred,1:nPred] <- sigGP^2*exp(-1/2*diffPP[1:nPred,1:nPred]/rho^2) + sigOE^2*IPP[1:nPred,1:nPred]
   SigPO[1:nPred,1:nObs ] <- sigGP^2*exp(-1/2*diffPO[1:nPred,1:nObs ]/rho^2)
   yObs[1:nObs] ~ dmnorm(mu[1:nObs], cov = SigOO[1:nObs,1:nObs])
})

constants <- list(nObs=nObs, nPred=nPred, diffOO=diffOO, diffPP=diffPP, diffPO=diffPO,
                  IOO=diag(nObs), IPP=diag(nPred), mu=rep(0,nObs))

data <- list(yObs = yObs)

inits <- list(rho = 1, sigGP = 1, sigOE = 1)

Rmodel <- nimbleModel(code, constants, data, inits)
```


### Create, compile, and run

```{r, message=FALSE}

## MCMC specification with no samplers
spec <- configureMCMC(Rmodel, nodes = NULL)

## this will be used for some checking, and setting up block sampler:
params <- Rmodel$getNodeNames(topOnly = TRUE)

## NOTE: the next line shouldn't work for you, since the MCMC API has changed.
## you can rebuild from source off nimble branch 'devel'.
## otherwise, using last public release of NIMBLE (0.3-1), use this line instead:
## spec$addSampler('RW_block', control = list(targetNodes = params))
spec$addSampler(params, 'RW_block')
```

We can debate about univariate vs. block sampling at some point.

```{r, message=FALSE}
## MCMC function
Rmcmc <- buildMCMC(spec)

## GP prediction function
## also uses the 'params' variable for specialization
Rpred <- gpPred(Rmodel, params)

## compile everything
Cmodel <- compileNimble(Rmodel)
Cmcmc  <- compileNimble(Rmcmc, project = Rmodel)
Cpred  <- compileNimble(Rpred, project = Rmodel)

## MCMC sampling
## 100,000 iterations
system.time(Cmcmc$run(100000))
```

About 25 seconds on my computer

That can be improved if necessary

```{r, message=FALSE}
samples <- as.matrix(Cmcmc$mvSamples)

## quick check
if(!identical(params, colnames(samples))) stop('problem')

## predict from GP model using posterior MCMC samples
system.time(Cpred$run(samples))
```

About 40 seconds on my computer

Again, could streamline the gpPred() function if necessary

```{r, message=FALSE}
## extract predictions: E and C
E <- Cpred$getE()
C <- Cpred$getC()
```



### Output

```{r}
E
```

```{r}
diag(C)
```

Black dots are the original 'xObs' and 'yObs'

Red dots are the predictions made at each 'xPred'

Red lines are plus/minus one standard error

```{r echo=FALSE, fig.width=8, fig.height=6}
plot(xObs, yObs, type='b', pch=19, xlim=range(c(xObs,xPred)), ylim=range(c(yObs,E)))
points(xPred, E, pch=20, col='red')
segments(x0=xPred, y0=E-sqrt(diag(C)), y1=E+sqrt(diag(C)), col='red')
```


