---
layout: post
category: ecology
tags:
  - early-warning
  - nimble
code: true
---


```{r settings}
library(knitr)
library(nimble)
library(sde)
library(ggplot2)
library(tidyr)
opts_chunk$set(dev='png', fig.width=5, fig.height=5, cache=1,results='hide')
```


Stick with SDE example as a reference point:

```{r data}
set.seed(123)
d <- expression(0.5 * (10-x))
s <- expression(1) 
data <- as.data.frame(sde.sim(X0=6,drift=d, sigma=s, T=20, N=100))
ggplot(data) + geom_line(aes(seq_along(x),x))
N <- length(data$x)
```

## LSN version ##

Modify the LSN model to explicitly model the changing parameter as a hidden, stochastic variable

```{r model, depends='detrend'}
lsn <- nimbleCode({
		theta ~ dunif(1e-10, 100.0)
		sigma_x ~ dunif(1e-10, 100.0)
		sigma_y ~ dunif(1e-10, 100.0)
					m ~ dunif(-1e2, 1e2)
			 x[1] ~ dunif(0, 100)
			 y[1] ~ dunif(0, 100) 

  for(i in 1:(N-1)){
    mu_x[i] <- x[i] + y[i] * (theta - x[i]) 
    x[i+1] ~ dnorm(mu_x[i], sd = sigma_x) 
    mu_y[i] <- y[i] + m * t[i]
    y[i+1] ~ dnorm(mu_y[i], sd = sigma_y) 
  }
})
```

Constants in the model definition are the length of the dataset, $N$ and the time points of the sample. Note we've made time explicit, we'll assume uniform spacing here. 

```{r constants, depends='data'}
constants <- list(N = N, t = 1:N)
```

Initial values for the parameters

```{r inits, depends='constants'}
inits <- list(theta = 6, m = 0, sigma_x = 1, sigma_y = 1, y = rep(1,N))
```


and here we go:

```{r compile, depends='model'}
Rmodel <- nimbleModel(code = lsn, 
                      constants = constants, 
                      data = data, 
                      inits = inits)
Cmodel <- compileNimble(Rmodel)
```



```{r spec, depends='compile', echo=TRUE, results = 'hide', message = FALSE}
mcmcspec <- configureMCMC(Rmodel, print=TRUE,thin=2e2)
Rmcmc <- buildMCMC(mcmcspec)
Cmcmc <- compileNimble(Rmcmc, project = Cmodel)
```


```{r run, depends='spec'} 
Cmcmc$run(1e6)
```


and examine results

```{r samples, depends='run'}
samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
dim(samples)
samples <- samples[,1:4]
long <- gather(samples)
```

```{r stats, depends='samples', results='markup'}
apply(samples, 2, mean)
```

```{r trace, depends='samples', fig.show='hold'}
ggplot(long) + 
	geom_line(aes(seq_along(value), value)) + 
	facet_wrap(~key, scale='free')
```

```{r histogram, depends='samples', fig.show='hold'}
ggplot(long) + 
	geom_density(aes(value)) + 
	facet_wrap(~key, scale='free')
```

```{r results='markup'}
sessionInfo()
```
