---
layout: post
category: computation
tags:
- rfishbase
- ropensci
- R
---

```{r include=FALSE}
library(knitr)
opts_chunk$set(results='hide')
```


Using `dplyr` calls on the back-end of the `rfishbase` re-write means working around the non-standard evaluation (NSE), as described in the [`dplyr` vignette]().

Grab the data I was using for this:

```{r}
library("dplyr")
downloader::download("https://github.com/cboettig/2015/raw/fc0d9185659e7976927d0ec91981912537ac6018/assets/data/2015-02-06-taxa.csv", "taxa.csv")
all_taxa <- read.csv("taxa.csv")
```

Consider a simple NSE `dplyr` call:

```{r}
x <- filter(all_taxa, Family == 'Scaridae')
```

The best SE version of this just needs to use the formula expression, `~`, the `_` SE version of the function and it's `.dots` argument:

```{r}
.dots <- list(~Family == 'Scaridae')
x1 <- filter_(all_taxa, .dots=.dots)

identical(x, x1)
```

This lets us treat the arguments (e.g. values of the factor on which we filter) as variables:


```{r}
family <- 'Scaridae'
.dots <- list(~Family == family)
x2 <- filter_(all_taxa, .dots=.dots)
identical(x, x2)
```  

If we want both the key and value to vary, we need to get pretty fancy to subvert the non-standard evaluation:

```{r}
library(lazyeval)
family <- 'Scaridae'
field <- 'Family'
.dots <- list(interp(~y == x, 
                     .values = list(y = as.name(field), x = family)))
x3 <- filter_(all_taxa, .dots=.dots)
identical(x, x3)
```

At bit more fun to wrap this into a function where we take arbitrary number of arguments as name-value pairs:

```{r}
query <- list(Family = 'Scaridae', SpecCode = 5537)
dots <- lapply(names(query), function(level){
    value <- query[[level]]
    interp(~y == x, 
                .values = list(y = as.name(level), x = value))
  })
  
x3 <-  filter_(all_taxa, .dots = dots) 
```
