---
published: false
layout: post
category: ecology
tags: 
- gp
- nimble

---



```{r}
knitr::opts_chunk$set(eval=FALSE)
library("gpmanagement")
library("tidyr")
library("dplyr")
library("ggplot2")
```


Programming around nimble: Can we break out a piece of the nimble model definition to be diefined by users, using the nimble format (e.g. the priors), while leaving some of the model to be specified seperately? 

At the moment this requires some trickery: http://stackoverflow.com/questions/31712646

It's interesting to observe just how much nimble's use (and extension!) of the powerful modeling expressions from BUGS has in common with the non-standard evaluation of Hadley's `dplyr` and friends.  With this in place in `gpmanagement` we can specify the prior externally.  

First, some sample data


```{r}  
# Should move my 'standard' sample data into the package?
  set.seed(1234)
  Tobs <- 40
  f <- function (x, h, p){
    sapply(x, function(x) {
        x <- pmax(0, x - h)
        x * exp(p[1] * (1 - x/p[2]) * (x - p[3])/p[2])
    })
  }
  p <- c(2, 100, 50)
  sigma_g <- 0.1
  z_g <- function() rlnorm(1, 0, sigma_g)
  x <- numeric(Tobs)
  x[1] <- 60
  for(t in 1:(Tobs-1))
    x[t+1] = z_g() * f(x[t], h=0, p=p)
  obs <- data.frame(x = c(0, 
                          pmax(rep(0,Tobs-1), x[1:(Tobs-1)])), 
                    y = c(0, 
                          x[2:Tobs]))
  xObs <- obs$x
  yObs <- obs$y
  xPred <- seq(0, 1.25 * max(xObs), length = 50)
```

```{r}
qplot(seq_along(x), x) + geom_line()
```


Now the GP estimation from NIMBLE.  Let's emphasize shorter length-scales with the prior to compare:

```{r}
# having this work with output from `nimbleCode` might be more natural than `expression`
priors <- expression({
  rho ~ dgamma(1, 10)
  sigGP ~ dunif(0, 1e5)
  sigOE ~ dunif(0, 1e5)
})
curve(dgamma(x, 1, 10))

```

```{r}
fit <- gp_setup(xObs[2:Tobs], yObs[2:Tobs], xPred)
```

```{r}
Cmcmc <- fit$Cmcmc 
Cpred <- fit$Cpred
Cmodel <- fit$Cmodel
```

```{r}
system.time(Cmcmc$run(100000))
```

```{r}
  samples <- as.matrix(Cmcmc$mvSamples)
  ## basic sanity check
  testthat::expect_identical(Cmodel$getNodeNames(topOnly = TRUE), colnames(samples))
```

predict from GP model using posterior MCMC samples

```{r}
system.time(Cpred$run(samples))
```


extract predictions: E and C


```{r}
  E <- Cpred$getE()
  C <- Cpred$getC()
  
obs <- data.frame(x = xObs, y = yObs)
pred <- data.frame(x = xPred, y = E, ymin = E - sqrt(diag(C)), ymax = E + sqrt(diag(C)))
ggplot2::ggplot(pred) + 
  geom_ribbon(aes(x = x,y = y, ymin = ymin, ymax = ymax), fill = "grey80") +
  geom_line(aes(x = x, y = y), size=1) + 
  geom_point(data = obs, aes(x,y)) +
  coord_cartesian(xlim = range(c(xObs, xPred)), ylim = range(c(yObs,E))) +
  theme_bw()

```


Posteriors

```{r}
samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
df <- tidyr::gather(samples)
ggplot(df) + 
  geom_density(aes(value)) + 
  facet_wrap(~key, scale='free')
```  


------------------------------------------

### Longer length scale



Now the GP estimation from NIMBLE.  Let's emphasize shorter length-scales with the prior to compare:

```{r}
# having this work with output from `nimbleCode` might be more natural than `expression`
priors <- expression({
  rho ~ dgamma(10, 1)
  sigGP ~ dunif(0, 1e5)
  sigOE ~ dunif(0, 1e5)
})
curve(dgamma(x, 10, 1))
```

```{r}
fit <- gp_setup(xObs[2:Tobs], yObs[2:Tobs], xPred)
```

```{r}
Cmcmc <- fit$Cmcmc 
Cpred <- fit$Cpred
Cmodel <- fit$Cmodel
```

```{r}
system.time(Cmcmc$run(100000))
```

```{r}
  samples <- as.matrix(Cmcmc$mvSamples)
  ## basic sanity check
  testthat::expect_identical(Cmodel$getNodeNames(topOnly = TRUE), colnames(samples))
```

predict from GP model using posterior MCMC samples

```{r}
system.time(Cpred$run(samples))
```


extract predictions: E and C


```{r}
  E2 <- Cpred$getE()
  C2 <- Cpred$getC()
```


sanity check: these should be different

```{r}
identical(E, E2)
identical(C, C2)
```

```{r}
obs <- data.frame(x = xObs, y = yObs)
pred <- data.frame(x = xPred, y = E2, ymin = E2 - sqrt(diag(C2)), ymax = E + sqrt(diag(C2)))
ggplot2::ggplot(pred) + 
  geom_ribbon(aes(x = x,y = y, ymin = ymin, ymax = ymax), fill = "grey80") +
  geom_line(aes(x = x, y = y), size=1) + 
  geom_point(data = obs, aes(x,y)) +
  coord_cartesian(xlim = range(c(xObs, xPred)), ylim = range(c(yObs,E))) +
  theme_bw()

```


Posteriors

```{r}
samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
df <- tidyr::gather(samples)
ggplot(df) + 
  geom_density(aes(value)) + 
  facet_wrap(~key, scale='free')
```  


