---
layout: post
category: ecology
tags: 
- gp
- nimble 

---

```{r }
library("nimble")
library("ggplot2")
library("tidyr")
library("dplyr")
```



## dmnorm example ##

from Daniel, who notes we need to be explicit about vector/matrix sizes on both LHS and RHS,

```{r}
code <- nimbleCode({
    mu ~ dunif(-100, 100)
    sigma ~ dunif(0, 100)
    rho ~ dunif(20, 100)
    muVec[1:N] <- mu * onesVector[1:N]
    Cov[1:N, 1:N] <- sigma^2 * exp(-dist[1:N, 1:N]/rho)
    g[1:N] ~ dmnorm(muVec[1:N], cov = Cov[1:N, 1:N])
    for (i in 1:N) {
        y[i] ~ dpois(exp(g[i]))
    }
})
constants <- list(N = 148, 
                  onesVector = rep(1, 148), 
                  dist = matrix(1, nrow=148, ncol=148))
y = rpois(148, 1) # vector of 148 observations (counts)
data <- list(y=y)

inits <- list(mu = 0, 
              sigma = 5, 
              rho = 60, 
              g = rep(0, 148))
m <- nimbleModel(code = code, constants = constants, data = data, inits = inits)
```




## GP Model, original specification ## 

This model definition doesn't work, since we do not define sizes explicitly



```{r warnings=TRUE, error=TRUE}
code <- nimbleCode({
   l ~ dunif(0,1e4)
   sigma.n ~ dunif(0, 1e4) 
   Sigma[1:N, 1:N] <- exp(-0.5 * diff / l ^ 2) + sigma.n ^ 2 * I
   y[1:N] ~ dmnorm(Mu, cov = Sigma)  
})
obs <- data.frame(x = c(-4, -3, -1,  0,  2),
                   y = c(-2,  0,  1,  2, -1))

N <- length(obs$x)
diff <- outer(obs$x, obs$x, function(xi, xj) (xi-xj)^2)
constants = list(N = N, x = obs$x, Mu = rep(0,N), diff = diff, I = diag(1,N))
inits <- list(l = 1, sigma.n = 10)
data <- data.frame(y = obs$y)
m <- nimbleModel(code = code, constants = constants, data = data, inits = inits)
```


## GP Model, v2 ## 

This model definition still doesn't work, even though it appears to be more explicit



```{r, error=TRUE}
code <- nimbleCode({

   l ~ dunif(0,100)
   sigma.n ~ dunif(0,100)
   SE <- function(xi,xj, l) exp(-0.5 * (xi - xj) ^ 2 / l ^ 2)
   Sigma[1:N, 1:N] <- outer(x[1:N], x[1:N], SE, l) + sigma.n ^ 2 * I[1:N, 1:N]
   y[1:N] ~ dmnorm(Mu[1:N], cov = Sigma[1:N, 1:N])  

})
obs <- data.frame(x = c(-4, -3, -1,  0,  2),
                   y = c(-2,  0,  1,  2, -1))
N <- length(obs$x)
constants = list(N = N, x = obs$x, Mu = rep(0,N), I = diag(1,N))
inits <- list(l = 1, sigma.n = 10)
data <- data.frame(y = obs$y)
nimbleModel(code = code, constants = constants, data = data, inits = inits)
```





## GP Model ##

A successful specification looks like this:

```{r}
code <- nimbleCode({

   l ~ dunif(0,1e5) # dgamma(5, 5)
   sigma.n ~ dunif(0, 1e5) # dgamma(5, 5) 
   Sigma[1:N, 1:N] <- exp(-0.5 * diff[1:N, 1:N] / l ^ 2) + sigma.n ^ 2 * I[1:N, 1:N]
   y[1:N] ~ dmnorm(Mu[1:N], cov = Sigma[1:N, 1:N])  

})
```


```{r}
obs <- data.frame(x = c(-4, -3, -1,  0,  2),
                   y = c(-2,  0,  1,  2, -1))

N <- length(obs$x)
diff <- outer(obs$x, obs$x, function(xi, xj) (xi-xj)^2)
constants = list(N = N, x = obs$x, Mu = rep(0,N), diff = diff, I = diag(1,N))
inits <- list(l = 1, sigma.n = 10)
data <- data.frame(y = obs$y)
m <- nimbleModel(code = code, constants = constants, data = data, inits = inits)
```


Define our mcmc procedure in Nimble

```{r }
my_mcmc <- function(code, constants, data, inits, n_iter=1e5, thin = 1e2){  
  Rmodel <- nimbleModel(code = code, constants = constants, data = data, inits = inits)
  Cmodel <- compileNimble(Rmodel)
  mcmcspec <- configureMCMC(Rmodel, print=FALSE,thin=thin)
  Rmcmc <- buildMCMC(mcmcspec)
  Cmcmc <- compileNimble(Rmcmc, project = Cmodel)
  Cmcmc$run(n_iter)
  samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
  #samples <- samples[,1:(length(inits)-1)]
  gather(samples)
}
```

```{r, message=FALSE}
df <- my_mcmc(code = code, constants = constants, data = data, inits = inits, n = 1e6)
```


#### Summary statistics

```{r , results='markup'}
summarise(group_by(df, key), mean=mean(value), std=sqrt(var(value)))
```

#### Traces

```{r , fig.show='hold'}
ggplot(sample_n(df,500)) + 
  geom_line(aes(seq_along(value), value)) + 
  facet_wrap(~key, scale='free')
```

#### Posteriors

```{r , fig.show='hold'}
ggplot(df) + 
  geom_density(aes(value)) + 
  facet_wrap(~key, scale='free')
```



## Time-series example

```{r}
f <- function (x, p){ x * exp(p[1] * (1 - x / p[2]) * (x - p[3]) / p[2]) }
p <- c(2, 8, 5)
z_g <- function() rlnorm(1, 0, 0.05)
T <- 40

set.seed(1234)
x <- numeric(T)
x[1] <- 5.5

for(t in 1:(T-1))
  x[t+1] = z_g() * f(x[t], p=p)

qplot(seq_along(x), x)
data <- data.frame(x=x)

```


#### model setup

```{r}
obs <- data.frame(x=x[1:(T-1)], y = x[2:T])
N <- length(obs$x)
diff <- outer(obs$x, obs$x, function(xi, xj) (xi-xj)^2)
constants = list(N = N, x = obs$x, Mu = rep(0,N), diff = diff, I = diag(1,N))
inits <- list(l = 1, sigma.n = 10)
data <- data.frame(y = obs$y)
```

#### Estimate GP bm MCMC with Nimble


```{r, message=FALSE}
df <- my_mcmc(code = code, constants = constants, data = data, inits = inits, n = 1e6)
```


#### Summary statistics

```{r , results='markup'}
summarise(group_by(df, key), mean=mean(value), std=sqrt(var(value)))
```

#### Traces

```{r , fig.show='hold'}
ggplot(sample_n(df,500)) + 
  geom_line(aes(seq_along(value), value)) + 
  facet_wrap(~key, scale='free')
```

#### Posteriors

```{r , fig.show='hold'}
ggplot(df) + 
  geom_density(aes(value)) + 
  facet_wrap(~key, scale='free')
```




