---
layout: post
category: ecology
tags: 
- early-warning
- regime-shifts

---

## Testing nimble method on Dai model

```{r}
devtools::install_github("cboettig/regimeshifts")
```


```{r include=FALSE, cache=FALSE}
library("nimble")
library("regimeshifts")
library("tidyr")
library("ggplot2")
```


Define our mcmc procedure in Nimble

```{r }
my_mcmc <- function(code, constants, data, inits, n_iter=1e5, thin = 1e2){
  Rmodel <- nimbleModel(code = code, constants = constants, data = data, inits = inits)
  Cmodel <- compileNimble(Rmodel)
  mcmcspec <- configureMCMC(Rmodel, print=FALSE,thin=thin)
  Rmcmc <- buildMCMC(mcmcspec)
  Cmcmc <- compileNimble(Rmcmc, project = Cmodel)
  Cmcmc$run(n_iter)
  samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
  samples <- samples[,1:(length(inits)-1)]
  gather(samples)
}
```



Generate the test data:

```{r}
set.seed(1000)
max_days <- 500
DF <- seq(0, 2000, length=max_days) # schedule for env degredation (increased dilution)
x <- numeric(max_days)   
x[1] <- 1.76e5 # initial density
   
for(day in 1:(max_days-1)){
 x[day+1] <- dai(x[day], DF = DF[day])
}

raw <- data.frame(t = 1:max_days, x = x)

```


Detrend: 

```{r detrend, depends='data'}
N <- length(raw$x)
raw$t <- 1:N
detrend <- loess(x ~ t, raw)
data <- data.frame(x = detrend$residuals/sqrt(var(detrend$residuals)))
qplot(raw$t, data$x, geom='line')
```

## OU Model ##

```{r}
ou <- nimbleCode({
   theta ~ dunif(1e-10, 100.0)
       r ~ dunif(1e-10, 20.0)
   sigma ~ dunif(1e-10, 100)
    x[1] ~ dunif(0, 100)

  for(t in 1:(N-1)){
    mu[t] <- x[t] + r * (theta - x[t]) 
    x[t+1] ~ dnorm(mu[t], sd = sigma) 
  }
})

ou_constants <- list(N = N)
ou_inits <- list(theta = 0, r = 1e-3, sigma = 1)
```

Run the mcmc

```{r message=FALSE}
df <- my_mcmc(code=ou, ou_constants, data, ou_inits)
```


#### Summary statistics

```{r , results='markup'}
summarise(group_by(df, key), mean=mean(value), std=sqrt(var(value)))
```

#### Traces

```{r , fig.show='hold'}
ggplot(df) + 
  geom_line(aes(seq_adf(value), value)) + 
  facet_wrap(~key, scale='free')
```

#### Posteriors

```{r , fig.show='hold'}
ggplot(df) + 
  geom_density(aes(value)) + 
  facet_wrap(~key, scale='free')
```






## LSN version ##

A modified version of the LSN model to explicitly model the changing parameter as a hidden variable changing at constant rate

```{r , depends='detrend'}
lsn <- nimbleCode({
   theta ~ dunif(-1e2, 1e2)
 sigma_x ~ dunif(1e-10, 1e2)
       m ~ dunif(-1e2, 1e2)
    x[1] ~ dunif(-1e3, 1e3)
    y[1] ~ dunif(-1e3, 1e3) 

  for(i in 1:(N-1)){
    mu_x[i] <- x[i] + y[i] * (theta - x[i]) 
    x[i+1] ~ dnorm(mu_x[i], sd = sigma_x) 
    y[i+1] <- y[i] + m * t[i] / t[N] 
    
  }
})

constants <- list(N = N, t = raw$t)
inits <- list(theta = 0, m = 0, sigma_x = 1, y = rep(1,N))
```


```{r}
df <- my_mcmc(code=lsn, constants, data, inits)
```

#### Summary statistics

```{r, results='markup'}
summarise(group_by(df, key), mean=mean(value), std=sqrt(var(value)))
```

#### Traces

```{r, fig.show='hold'}
ggplot(long) + 
  geom_line(aes(seq_along(value), value)) + 
  facet_wrap(~key, scale='free')
```

#### Posteriors

```{r, fig.show='hold'}
ggplot(long) + 
	geom_density(aes(value)) + 
	facet_wrap(~key, scale='free')
```




## LSN, stochastic hidden variable ##

#### Define and model and run MCMC


```{r }
lsn <- nimbleCode({
   theta ~ dunif(-1e2, 1e2)
   sigma_x ~ dunif(1e-10, 1e2)
   sigma_y ~ dunif(1e-10, 1e2)
       m ~ dunif(-1e2, 1e2)
    x[1] ~ dunif(-1e3, 1e3)
    y[1] ~ dunif(-1e3, 1e3)

  for(i in 1:(N-1)){
    mu_x[i] <- x[i] + y[i] * (theta - x[i])
    x[i+1] ~ dnorm(mu_x[i], sd = sigma_x)
    mu_y[i] <- y[i] + m * t[i] / t[N]
    y[i+1] ~ dnorm(mu_y[i], sd = sigma_y) 
  }
})

constants <- list(N = N, t = raw$t)
inits <- list(theta = 0, m = 0, sigma_x = 1, sigma_y = 1, y = rep(1,N))
```



```{r}
df <- my_mcmc(code=lsn, constants, data, inits)
```


#### Summary statistics


```{r, results='markup'}
summarise(group_by(df, key), mean=mean(value), std=sqrt(var(value)))
```


#### Traces

```{r, fig.show='hold'}
ggplot(df) + 
  geom_line(aes(seq_along(value), value)) + 
  facet_wrap(~key, scale='free')
```

#### Posteriors 

```{r, fig.show='hold'}
ggplot(df) + 
  geom_density(aes(value)) + 
	facet_wrap(~key, scale='free')
```


--------------


```{r results='markup'}
sessionInfo()
```



