---
layout: post
category: ecology
tags: 
- gp
- nimble

---

Original go at nimble gp model, but this requires rather manual manipulation of parameter posteriors to get the GP posteriors.  


```{r }
library("MASS")
library("mcmc")
library("nimble")
library("ggplot2")
library("tidyr")
library("dplyr")
library("nonparametricbayes")
library("pdgControl")
```

## sample data ##

```{r}
set.seed(1234)
Tobs <- 40
f <- RickerAllee
sigma_g <- 0.05
z_g <- function() rlnorm(1, 0, sigma_g)
p <- c(2, 8, 5)
```


```{r }
x_grid <- seq(0, 15, length=50)
h_grid <- x_grid
profit <- function(x,h) pmin(x, h)
delta <- 0.01
OptTime <- 50  # stationarity with unstable models is tricky thing
xT <- 0    # terminal condition
x0 <- 10   # simulation under policy starts from
MaxT <- 1000 # timeout for value iteration convergence
```


```{r}
x <- numeric(Tobs)
x[1] <- 5.5
nz <- 1
for(t in 1:(Tobs-1))
  x[t+1] = z_g() * f(x[t], h=0, p=p)
obs <- data.frame(x = c(rep(0,nz), 
                        pmax(rep(0,Tobs-1), x[1:(Tobs-1)])), 
                  y = c(rep(0,nz), 
                        x[2:Tobs]))
ggplot(data.frame(time = 1:Tobs, x=x), aes(time,x)) + geom_line()
```

## GP Model ##


```{r}
code <- nimbleCode({

   l ~ dgamma(10, 1) 
   sigma.n ~ dunif(0, 1e5)
   sigma.k  ~ dunif(0, 1e5)
   Sigma[1:N, 1:N] <- sigma.k ^ 2 * exp(-0.5 * diff[1:N, 1:N] / l ^ 2) + 
                      sigma.n ^ 2 * I[1:N, 1:N]
   y[1:N] ~ dmnorm(Mu[1:N], cov = Sigma[1:N, 1:N])  

})
```


```{r}
N <- length(obs$x)
diff <- outer(obs$x, obs$x, function(xi, xj) (xi-xj)^2)
constants = list(N = N, x = obs$x, Mu = rep(0,N), diff = diff, I = diag(1,N))
inits <- list(l = 1, sigma.n = 10, sigma.k = 1)
data <- data.frame(y = obs$y)


```

```{r}
Rmodel <- nimbleModel(code = code, constants = constants, data = data, inits = inits)
```


Define our mcmc procedure in Nimble

```{r }
Cmodel <- compileNimble(Rmodel)
mcmcspec <- configureMCMC(Rmodel, print=FALSE)
Rmcmc <- buildMCMC(mcmcspec)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

```{r}
system.time(
Cmcmc$run(1e6)
)
```

```{r}
samples <- as.data.frame(as.matrix(Cmcmc$mvSamples))
df <- gather(samples)
```


#### Posteriors

```{r , fig.show='hold'}
ggplot(df) + 
  geom_density(aes(value)) + 
  facet_wrap(~key, scale='free') + 
  scale_x_log10()
```

## Calculating transition probabilities

Several ways we might go about this.

Technically all policies could be updated in response to new information, though this requires repeating the estimation process, or at least a Bayesian updating step (e.g. passive learning).  Typically these steps are separated into estimating the policy and then implementing the policy separately, so we will focus on this case.

The transition probability is conditional 




```{r}
## Should be able to extract this from nimble...
predict <- 
function (posteriors, obs, x_predict) {
    out <- lapply(data.frame(t(posteriors)), 
      function(sample) {
      
        l <- sample[1]
        sigma.k <- sample[2]
        sigma.n <- sample[3]
       
        
        SE <- function (Xi, Xj, l) sigma.k * exp(- 0.5 * (Xi - Xj) ^ 2 / l ^ 2)
        cov <- function(X, Y) outer(X, Y, SE, l)
            
        cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
        Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
        Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x) %*% 
            cov_xx_inv %*% cov(obs$x, x_predict)
        list(Ef = Ef, Cf = Cf, Vf = diag(Cf))
    })
    Ef_posterior <- sapply(out, `[[`, "Ef")
    Cf_posterior <- sapply(out, `[[`, "Cf")
    Vf_posterior <- sapply(out, `[[`, "Vf")
    E_Ef <- rowMeans(Ef_posterior)
    E_Cf <- matrix(apply(Cf_posterior, 1, sum)/dim(Cf_posterior)[2], 
        ncol = sqrt(dim(Cf_posterior)[1]))
    E_Vf <- diag(E_Cf)
    Cf_posterior <- lapply(out, `[[`, "Cf")
    list(Ef_posterior = Ef_posterior, Vf_posterior = Vf_posterior, 
        Cf_posterior = Cf_posterior, E_Ef = E_Ef, E_Cf = E_Cf, 
        E_Vf = E_Vf)
}

gp_posterior <- predict(sample_n(samples, 100), obs, x_grid)
```



```{r}
matrices_gp <- gp_transition_matrix(gp_posteriort$Ef_posterior, gp_posterior$Vf_posterior, x_grid, h_grid) 
opt_gp <- value_iteration(matrices_gp, x_grid, h_grid, MaxT, xT, profit, delta, reward)
```

--------------------------------------------

## Comparison

```{r}
s2.p <- c(5,5)  
d.p = c(10, 1/0.1)
gp <- gp_mcmc(obs$x, y=obs$y, n=1e5, s2.p = s2.p, d.p = d.p)
gp_dat <- gp_predict(gp, x_grid, burnin=1e4, thin=300)
```



```{r gp-opt, dependson="gp-output"}
matrices_gp <- gp_transition_matrix(gp_dat$Ef_posterior, gp_dat$Vf_posterior, x_grid, h_grid) 
opt_gp <- value_iteration(matrices_gp, x_grid, h_grid, MaxT, xT, profit, delta)
```

```{r mle-opt, dependson="mle-output"}
matrices_true <- f_transition_matrix(f, p, x_grid, h_grid, sigma_g)
opt_true <- value_iteration(matrices_true, x_grid, h_grid, OptTime=MaxT, xT, profit, delta)
```




```{r}
policy <- list(GP = opt_gp$D, exact = opt_true$D)
sets <- expand.grid(reps = 1:100, model = c("exact", "GP")) %>% group_by(reps, model)

sim_fn <- function(df){
  set.seed(df$reps)
  ForwardSimulate(f, p, x_grid, h_grid, x0, 
                  D = policy[[as.character(df$model)]], 
                  z_g, profit=profit, OptTime = OptTime)
  
}
sims <- sets %>% do(sim_fn(.))
```

```{r}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
ggplot(sims) + 
  geom_line(aes(time, fishstock, group=interaction(reps,model), col=model), alpha=.1) + 
  facet_wrap(~model) +
  scale_colour_manual(values=colorkey, guide=FALSE)
```

```{r}

policies_plot <- function(policy){
  policy_df <- 
    data.frame(model = c("GP", "exact"))  %>% 
    group_by(model) %>% 
    do(data.frame(stock = x_grid, 
                  escapement = x_grid - h_grid[policy[[as.character(.$model)]]]))

  ggplot(policy_df, aes(stock, escapement, color=model)) +
    geom_line() + 
    facet_wrap(~model) +
    xlab("stock size, x(t)") + 
    ylab("escapement, S(t)")  +
    scale_colour_manual(values=colorkey, guide=FALSE)
}

policies_plot(policy)

```

```{r}
sims %>% mutate(net_profit = sum(profit)) -> sims
```


