---
layout: post
category: ecology
published: false
---



Load libraries and data as before

```{r}
library("ggplot2")
library("dplyr")
library("tidyr")
library("knitcitations")
library("rgpdd")
library("FKF")
```


```{r}
gpdd_data %>% 
  filter(MainID %in% filtered$MainID) %>%
  select(MainID, Population, SampleYear) %>%
  group_by(MainID) %>% 
  mutate(logN = log(Population)) ->
df
```

```{r}
gpdd_main %>% 
  filter(SamplingProtocol == "Count",
         SourceDimension %in% c("Count", "Index"), 
         SamplingFrequency == "1",
         DatasetLength >= 15) %>%
  select(MainID) %>%
  arrange(MainID) ->
filtered
```

Define functions for each model fit that take the data and return optim output (inc estimated parameters, minus log likelihood


```{r}

fit_ssg <- function(y){
    o <- optim(c(dt = mean(y), Tt = 1, HHt = var(y)/2, GGt = var(y)/2),
                 fn =  function(par, ...)
                   -fkf(dt = matrix(par[1]),
                        Tt = matrix(par[2]),
                        HHt = matrix(par[3]), 
                        GGt = matrix(par[4]), 
                        ...)$logLik,   
                 a0 = y[1], 
                 P0 = matrix(10), 
                 ct = matrix(0),
                 Zt = matrix(1), 
                 yt = rbind(y), 
                 check.input = FALSE, 
                 ...)
  c(o, list(a0 = y[1], n = length(y)))
}

fit_ssrw <- function(y){
    o <- optim(c(dt=mean(y), HHt = var(y)/2, GGt = var(y)/2),
                 fn =  function(par, ...)
                   -fkf(dt = matrix(par[1]), HHt = matrix(par[2]), 
                        GGt = matrix(par[3]), ...)$logLik,   
                 a0 = y[1], P0 = matrix(10), ct = matrix(0), Tt = matrix(1),
                 Zt = matrix(1), yt = rbind(y), check.input = FALSE, 
                 ...)
  c(o, list(a0 = y[1], n = length(y)))
}

fit_g <- function(y){
  o <- optim(c(dt = mean(y), Tt=1, HHt = var(y)),
                 fn =  function(par, ...)
                   -fkf(dt = matrix(par[1]), Tt = matrix(par[2]), 
                        HHt = matrix(par[3]), ...)$logLik,   
                 a0 = y[1], P0 = matrix(10), ct = matrix(0), GGt = matrix(0),
                 Zt = matrix(1), yt = rbind(y), check.input = FALSE, 
                 ...)
  c(o, list(a0 = y[1], n = length(y)))
}

fit_rw <- function(y){
 o <-  optim(c(dt=mean(y), HHt = var(y)),
                 fn =  function(par, ...)
                   -fkf(dt = matrix(par[1]), HHt = matrix(par[2]), ...)$logLik,   
                 a0 = y[1], P0 = matrix(10), ct = matrix(0), Tt = matrix(1),
                 GGt = matrix(0), Zt = matrix(1), yt = rbind(y), check.input = FALSE, 
                 ...)
  c(o, list(a0 = y[1], n = length(y)))
}

```


## Simulating, bootstrapping

FKF package doesn't bother to define a simulation method, so we can simply define one directly from the state equations; though a C implementation might be preferred. We will ignore the multi-variate definition for simplicity here.

```{r}

use <- function(vec, x, default){
  if(! x %in% names(vec))
    default
  else
    vec[[x]]
}

sim.fkf <- function(fit){
  n <- fit$n
  dt <- fit$par[["dt"]]
  HHt <- fit$par[["HHt"]]
  Tt <- use(fit$par, "Tt", 1)
  GGt <- use(fit$par, "GGt", 0)
  a0 <- fit$a0
  ct <- 0
  Zt <- 1
  
  a <- numeric(n)
  y <- numeric(n)
  eta <- rnorm(n, dt, sqrt(HHt))
  epsilon <- rnorm(n, ct, sqrt(GGt))
  a[1] <- a0
		for(t in 1:(n-1)){
		  a[t+1] <- Tt * a[t] + eta[t]
		  y[t] <- Zt * a[t] + epsilon[t]
		}
		y[n] <- Zt * a[n] + epsilon[n]
  y
}

```

With fitting and simulating functions in place, defining the bootstrap is straight forward. We define these separately for the state-space Gompertz (ssg; i.e. the model with both density dependence and observational errors) and the Gompertz (g; density dependence, no observational error).  We compare in each case to the simulations of the corresponding model without density dependence.  


```{r}
bootstap_ssg <- function(df, N=100){
  ssg <- fit_ssg(df$logN)
  ssrw <- fit_ssrw(df$logN)
  sims <- replicate(N, sim_fkf(ssrw)) 
  null <- sapply(sims, function(y) fit_ssrw(y)$value)
  
  lr <- 2 * (ssrw$value - ssg$value)
  null_dist <- 2 * (null - ssg$value) 
  sum(which(null_dist < lr))/N
}
```

```{r}
bootstap_g <- function(df, N=100){
  g <- fit_g(df$logN)
  rw <- fit_rw(df$logN)
  sims <- replicate(N, sim_fkf(rw)) 
  null <- sapply(sims, function(y) fit_rw(y)$value)
  
  lr <- 2 * (rw$value - g$value)
  null_dist <- 2 * (null - g$value) 
  sum(which(null_dist < lr))/N
}

  
```

With these functions defined, we can perform the actual analysis:

```{r}
df %>% group_by(MainID) %>% do(bootstrap_ssg(.)) -> ssg_p_values
```


```{r}
df %>% group_by(MainID) %>% do(bootstrap_g(.)) -> ssg_p_values
```


The study also creates simulated datasets based on the real data but explicitly making the assumption of either density independence (DI) or density dependence (DD).  For each dataset, a density-independent simulated dataset is created by simulating under the SSRW model that was fit. The density-dependent model is created by explicitly fixing the density dependent parameter ($c$ in the language of the paper, `Tt` in FKF notation) to 0.8 and estimating the other parameters of this modified SSG model:

```{r}

fit_dd <- function(y){
    o <- optim(c(dt = mean(y), Tt = 1, HHt = var(y)/2, GGt = var(y)/2),
                 fn =  function(par, ...)
                   -fkf(dt = matrix(par[1]),
                        HHt = matrix(par[2]), 
                        GGt = matrix(par[3]), 
                        ...)$logLik,
                 Tt = matrix(0.8),
                 a0 = y[1], 
                 P0 = matrix(10), 
                 ct = matrix(0),
                 Zt = matrix(1), 
                 yt = rbind(y), 
                 check.input = FALSE, 
                 ...)
    
  c(list(par = c(dt=o$par[[1]], Tt = 0.8, HHt = o$par[[2]], GGt = o$par[[3]]),
         value = o$value),
  list(a0 = y[1], n = length(y)))
}

```



```{r}

df %>% group_by(MainID) %>% sim_fkf(fit_ssrw(.$logN)) -> DI
df %>% group_by(MainID) %>% sim_fkf(fit_dd(.$logN)) -> DD


```

